Fractional Types

### 2019-11-21 fracGC/p.tex
done with space section for now

\noindent where the operations take an implicit argument asserting the
given type is non-empty. The operational semantics of the language
interprets each combinator (except $\eta$ and $\epsilon$) of type
$t_1 \leftrightarrow t_2$ as a (reversible) function mapping values of
type $t_1$ to values of type $t_2$ (and vice-versa). Extending this
interpreter to handle $\eta$ and $\epsilon$ requires at this point
that the interpreter returns either a value of the appropriate or an
exception:

@@

%%%%%
\subsection{Examples}

@@

If a type has size $n$ we will use a number $0..(n-1)$ to represent
the values of the types. In binary, this would use $\log{n}$ bits.

The abstract machine has a register representing the code and a
pointer. If the code is acting on a type with $n$ values, the pointer
has $\log{n}$ bits (rounded).

In plain pi, if the input is representable using $n$ bits then this is
the only memory you get to use throughout the entire circuit. The
circuit just transforms these $n$ bits; example.

Use Fin size as proxy for memory size

@@

@@

CNOT trace in plain pi:

@@

@@

In frac pi, you are however allowed to use more than $n$ bits in the
intermediate steps as eta is used to allocate more space.

@@

zigzag trace

>>>>

\noindent where the operations take an implicit argument asserting the
given type is non-empty. A \emph{big step operational semantics} for
the language interprets each combinator (except $\eta$ and $\epsilon$)
of type $t_1 \leftrightarrow t_2$ as a (reversible) function mapping
values of type $t_1$ to values of type $t_2$ (and
vice-versa).\footnote{In Sec.~\ref{sec:space} we will define a small
  step abstract machine that exposes the space used at each step.}
Extending this interpreter to handle $\eta$ and $\epsilon$ requires
that the interpreter returns either a value of the appropriate or an
exception:

@@

%%%%%
\subsection{Examples}
\label{sub:id}

@@

Before before the abstract machine for the full system with
fractionals, we introduce some of the common technical details in the
simpler setting of the core $\Pi$ language. First two simple facts
about cardinalities:

@@

The cardinalities of the types (other than the fractional type) map
directly to the natural numbers. An easy proposition to verify is that
every combinator must be between types of the same cardinalities;
indeed one of the standard interpretations of $\Pi$ combinators is
that they are permutations on finite sets, i.e., they just shuffle
elements around without changing the sizes of the underlying sets.

To make the space usage explicit in every step of evaluation, we
define an abstract machine whose states are pairs of a code register
and a memory register. In a low-level implementation the code register
would contain a pointer to the current instruction and the memory
register would contain a bit-level runtime representation of values
with one bit for boolean values, etc. In the Agda formalization, we
include the current combinator in the code register, and represent the
memory register using an index into an enumeration. Thus a memory
register containing a boolean value would be represented in Agda as 𝔽
::  𝕋 :: [] [ Fin.0F ] or 𝔽 :: 𝕋 :: [] [ Fin.1F ] depending on whether
the values is \textsf{false} or \textsf{true}. The prepending list
contains all possible values of the type and the number in between [\ldots ] 
is an index into the list, representing a finite number that
is guaranteed to be in bounds. In a low-level runtime representation
only a binary encoding of this finite would need to be stored but in
order to just prove that space is preserved, we can just inspect the
size of the prepended list. Putting it all together, machine states and
small-step reductions are formalized as follows:

@@

The constructor for machine states takes a number which is the size of
the memory list. This size is maintained throughout the execution,
even if the intermediate types change. As a small example, the
following is a trace of the execution of \AgdaFunction{CNOT} applied
to 𝕋, 𝕋:

@@

Note how the size of the list remains equal to 4 throughout the
execution, even if the individual elements change as types change and
the current index (i.e., the current value) changes.

@@

Using the same assumptions as before, machine states and small step
reductions are formalized as follows:

@@

The main difference is that now the state constructor does not take a
fixed size and indeed the size changes during execution as appropriate
to fit the elements of the type $A$. In particular, since fractional
values have a trivial representation, the size of a fractional value
is 1, and hence $\eta$ used at type $A$ changes the size of the memory
list by the size of $A$. As an illustrative example, we include a
trace of the execution of \AgdaFunction{id'} from Sec.~\ref{sub:id}
applied to 𝕋:
